cola Report for Consensus Partitioning (SD:skmeans)
==================

**Date**: `r format(as.POSIXlt(Sys.time()), usetz = TRUE)`, **cola version**: `r installed.packages()["cola", "Version"]`

----------------------------------------------------------------

<style type='text/css'>

body, td, th {
   font-family: Arial,Helvetica,sans-serif;
   background-color: white;
   font-size: 13px;
  max-width: 800px;
  margin: auto;
  margin-left:210px;
  padding: 0px 10px 0px 10px;
  border-left: 1px solid #EEEEEE;
  line-height: 150%;
}

tt, code, pre {
   font-family: 'DejaVu Sans Mono', 'Droid Sans Mono', 'Lucida Console', Consolas, Monaco, 

monospace;
}

h1 {
   font-size:2.2em;
   line-height:150%;
}

h2 {
   font-size:1.8em;
}

h3 {
   font-size:1.4em;
}

h4 {
   font-size:1.0em;
}

h5 {
   font-size:0.9em;
}

h6 {
   font-size:0.8em;
}

a {
  text-decoration: none;
  color: #0366d6;
}

a:hover {
  text-decoration: underline;
}

a:visited {
   color: #0366d6;
}

pre, img {
  max-width: 100%;
}
pre {
  overflow-x: auto;
}
pre code {
   display: block; padding: 0.5em;
}

code {
  font-size: 92%;
  border: 1px solid #ccc;
}

code[class] {
  background-color: #F8F8F8;
}

table, td, th {
  border: 1px solid #ccc;
}

blockquote {
   color:#666666;
   margin:0;
   padding-left: 1em;
   border-left: 0.5em #EEE solid;
}

hr {
   height: 0px;
   border-bottom: none;
   border-top-width: thin;
   border-top-style: dotted;
   border-top-color: #999999;
}

@media print {
   * {
      background: transparent !important;
      color: black !important;
      filter:none !important;
      -ms-filter: none !important;
   }

   body {
      font-size:12pt;
      max-width:100%;
   }

   a, a:visited {
      text-decoration: underline;
   }

   hr {
      visibility: hidden;
      page-break-before: always;
   }

   pre, blockquote {
      padding-right: 1em;
      page-break-inside: avoid;
   }

   tr, img {
      page-break-inside: avoid;
   }

   img {
      max-width: 100% !important;
   }

   @page :left {
      margin: 15mm 20mm 15mm 10mm;
   }

   @page :right {
      margin: 15mm 10mm 15mm 20mm;
   }

   p, h2, h3 {
      orphans: 3; widows: 3;
   }

   h2, h3 {
      page-break-after: avoid;
   }
}
</style>

```{r, echo = FALSE, message = NA}
library(knitr)
library(markdown)
options(width = 100)
options(digits = 3)
opts_chunk$set(
    comment = "#>",
    fig.path = "figure_cola/",
    warning = FALSE,
    message = FALSE
)
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(genefilter))
ht_opt$message = FALSE

n_cores = get_nc(cores)
```


## Summary



First the variable is renamed to `res`.

```{r, eval = FALSE, echo = TRUE}
res = cola.out["SD:skmeans"]
```

```{r, echo = FALSE}
res = object
```

Dimension of the input matrix:

```{r}
mat = get_matrix(res)
dim(mat)
```

Top-value method used:

```{r}
res@top_value_method
```

Partitioning method used:

```{r}
res@partition_method
```

The density distribution for each sample is visualized as in one column in the
following heatmap. The clustering is based on the distance which is the
Kolmogorov-Smirnov statistic between two distributions.

```{r, echo = FALSE, message = NA}
t1 = Sys.time()
message(qq("* making density heatmap of global distribution for each sample (@{ncol(mat)} samples)"), appendLF = FALSE)
```

```{r density-heatmap}
library(ComplexHeatmap)
densityHeatmap(mat, top_annotation = HeatmapAnnotation(df = get_anno(res), 
    col = get_anno_col(res)), ylab = "value", cluster_columns = TRUE, show_column_names = FALSE,
    mc.cores = 1)
```

```{r, echo = FALSE, message = NA}
t2 = Sys.time(); message(paste0(", ", gsub("^ +", "", format(t2 - t1))))
```

 
## Results

```{r, results = "asis", echo = FALSE}
fs = min(c(3*1, 14))/1

is_hc = FALSE
node = NULL
single_report = TRUE

all_k = res@k

i_section = 1
n_section = 1
top_value_method = res@top_value_method
partition_method = res@partition_method

tmpfile = tempfile(tmpdir = cola:::.ENV$TEMP_DIR)

brew(file.path(cola:::TEMPLATE_DIR, "section_template.Rmd-template"), tmpfile)
rmd = paste(readLines(tmpfile), collapse = "\n")
invisible(file.remove(tmpfile))

ComplexHeatmap:::dev.null()
txt = knit(text = rmd, quiet = TRUE)
devlt = dev.list()
if(names(devlt)[1] == "quartz_off_screen") {
	invisible(dev.off())
}
invisible(dev.off())
cat(txt, "\n")
```

## Session info

```{r}
sessionInfo()
```

```{r, echo = FALSE, message = NA}
message("-----------------------------------------------------------------")
```


<script type="text/javascript">
$(function() {
    $("#TOC > ul > li > a").remove();
}); 
</script>
